PineAPPL:\\ combining EW and QCD corrections for PDF fits

### 2020-08-21 paper/app-pineappl.tex
Improve PineAPPL appendix
- move installation appendix towards the end
- include complete code example

\textsc{PineAPPL} currently consists of three parts: 1) the library itself, which is a dependency for the other parts, 2) the helper program \texttt{pineappl}, which allows one to read \textsc{PineAPPL} grids from the command line and make predictions with it, and finally 3) the C interface, which is intended to be used in Monte Carlo integrators to generate the grids.

After being installed, one can compile and link against the library.
See appendix~\ref{app:example-program} for an example.
Updated installation instructions are kept in the file \texttt{README.md} in \textsc{PineAPPL}'s repository at \url{https://github.com/N3PDF/pineappl/}.

@@

The program \texttt{pineappl} can be used to perform quick convolutions and other calculations with existing grids on the command line.
If started without any arguments, it prints its help and lists all supported subcommands:
\begin{verbatim}
 $ pineappl
pineappl 0.2.0
Read, write, and query PineAPPL grids

@@

\paragraph{Channels and Luminosity function.}
Sometimes it is useful to know which partons contribute the most and by how much.
This is what the subcommand \texttt{channels} shows:
\begin{verbatim}
 $ pineappl channels ATLASZHIGHMASS49FB.pineappl \
 > NNPDF31_nlo_as_0118_luxqed --limit 5
bin xmin xmax lumi  size
---+----+----+----+------+---+------+---+------+---+------+---+-----
  0  116  130  #15 27.42%  #0 27.42%  #5 24.52% #20 24.51% #30 1.30%
  1  130  150  #15 28.83%  #0 28.83%  #5 21.78% #20 21.78% #30 1.86%
  2  150  170  #15 29.87%  #0 29.82%  #5 19.63% #20 19.62% #30 2.32%
  3  170  190  #15 30.54%  #0 30.53% #20 18.17%  #5 18.13% #30 2.58%
  4  190  210   #0 31.06% #15 31.04%  #5 17.07% #20 17.04% #30 2.75%
  5  210  230   #0 31.46% #15 31.38% #20 16.27%  #5 16.24% #30 2.86%
  6  230  250   #0 31.73% #15 31.64% #20 15.66%  #5 15.62% #30 2.91%
  7  250  300   #0 32.07% #15 32.02%  #5 14.91% #20 14.86% #30 2.98%
  8  300  400   #0 32.62% #15 32.56%  #5 13.83% #20 13.81% #30 2.98%
  9  400  500  #15 33.14%  #0 33.12% #20 12.89%  #5 12.87% #30 2.93%
 10  500  700   #0 33.71% #15 33.31% #20 12.28%  #5 12.26% #30 2.93%
 11  700 1000  #15 34.08%  #0 33.71% #20 11.59%  #5 11.51% #30 3.06%
 12 1000 1500   #0 33.95% #15 33.88% #20 11.12%  #5 10.88% #30 3.56%
\end{verbatim}
The first three columns are known from \texttt{convolute}.
The next columns (the switch \texttt{-{}-limit 5} limits the output to five columns) show first the channel index and then the relative size of the corresponding contribution.
Since the contribution of a partonic channel can be negative, the columns are sorted ignoring the sign of the contribution.
The first line shows that for bin 0, i.e.\ for the range $\SI{116}{\giga\electronvolt} < M_{\ell \bar{\ell}} < \SI{130}{\giga\electronvolt}$, the cross section is dominated by partonic channel \texttt{\#15} (\SI{27.42}{\percent}), following by partonic channel \texttt{\#0} with same size, then channel \texttt{\#5}, etc.
The meaning of the channel numbers is given by using the subcommand \texttt{luminosity} (only an excerpt is shown):
\begin{verbatim}
id    entry
--+------------+------------+------------
0  1 × ( 2, -2) 1 × ( 4, -4)
5  1 × ( 1, -1) 1 × ( 3, -3)
15 1 × (-4,  4) 1 × (-2,  2)
20 1 × (-3,  3) 1 × (-1,  1)
30 1 × (22, 22)
\end{verbatim}
This shows that channel \texttt{\#0} represents the up-type quark--anti-quark contributions (shown with PDG id 2 and 4 for up and charm quarks, which have the same matrix elements), channel \texttt{\#15} is the same channel with its initial states transposed, channels \texttt{\#5} and \texttt{\#20} are the down-type quark--anti-quark channels, and channel \texttt{\#30} is the photon--photon channel.
The size of the remaining channels is smaller than the photon--photon channel.
The factors \texttt{1} are not important here, but in general they can contain CKM values and charge factors that, if kept in the squared matrix elements, would not allow for sharing a single matrix element for different quark flavours and therefore slow down the calculation. A complete list of the most important channels and of their
contribution to the cross section for all of the processes discussed in
section~\ref{sec:results} is collected in appendix~\ref{app:lumis}.

@@

\subsection{Example Monte Carlo program}
\label{app:example-program}

@@

The following listing shows how to setup \textsc{PineAPPL} using its C interface in a simple Monte Carlo integrator for calculating the double-photon contribution to Drell--Yan lepton-pair production at the LHC.
All \textsc{PineAPPL} functions have the prefix \texttt{pineappl\_}.
The full example can be found at \url{https://github.com/N3PDF/pineappl/tree/master/examples/capi-dy-aa} and the full documentation of the C API can be found at \url{https://docs.rs/pineappl_capi}.

\begin{lstlisting}[language=C++,mathescape=true]
// create a new luminosity function for the $\gamma\gamma$ initial state
auto* lumi = pineappl_lumi_new();
int32_t pdg_ids[] = { 22, 22 };
double ckm_factors[] = { 1.0 };
pineappl_lumi_add(lumi, 1, pdg_ids, ckm_factors);

@@

// destroy the object
pineappl_grid_delete(grid);
\end{lstlisting}

>>>>

\textsc{PineAPPL} currently consists of three parts: 1) the library itself, which is a dependency for the other parts, 2) the helper program \texttt{pineappl}, which allows one to read \textsc{PineAPPL} grids from the command line and make predictions with it (explained in section~\ref{app:pineappl-demo}), and finally 3) the C interface, which is intended to be used in Monte Carlo integrators to generate the grids.

@@

The program \texttt{pineappl} can be used to perform quick convolutions and other calculations with existing grids on the command line.
If started without any arguments, it prints its help and lists all supported subcommands:
\begin{verbatim}
 $ pineappl
pineappl 0.3.0
Read, write, and query PineAPPL grids

@@

\paragraph{Channels and Luminosity function.}
Sometimes it is useful to know which partons contribute the most and by how much.
This is what the subcommand \texttt{channels} shows:
\begin{verbatim}
 $ pineappl channels ATLASZHIGHMASS49FB.pineappl \
 > NNPDF31_nlo_as_0118_luxqed --limit 5
bin xmin xmax lumi  size
---+----+----+----+------+---+------+---+------+---+------+---+-----
  0  116  130  #15 27.42%  #0 27.42%  #5 24.52% #20 24.51% #30 1.30%
  1  130  150  #15 28.83%  #0 28.83%  #5 21.78% #20 21.78% #30 1.86%
  2  150  170  #15 29.87%  #0 29.82%  #5 19.63% #20 19.62% #30 2.32%
  3  170  190  #15 30.54%  #0 30.53% #20 18.17%  #5 18.13% #30 2.58%
  4  190  210   #0 31.06% #15 31.04%  #5 17.07% #20 17.04% #30 2.75%
  5  210  230   #0 31.46% #15 31.38% #20 16.27%  #5 16.24% #30 2.86%
  6  230  250   #0 31.73% #15 31.64% #20 15.66%  #5 15.62% #30 2.91%
  7  250  300   #0 32.07% #15 32.02%  #5 14.91% #20 14.86% #30 2.98%
  8  300  400   #0 32.62% #15 32.56%  #5 13.83% #20 13.81% #30 2.98%
  9  400  500  #15 33.14%  #0 33.12% #20 12.89%  #5 12.87% #30 2.93%
 10  500  700   #0 33.71% #15 33.31% #20 12.28%  #5 12.26% #30 2.93%
 11  700 1000  #15 34.08%  #0 33.71% #20 11.59%  #5 11.51% #30 3.06%
 12 1000 1500   #0 33.95% #15 33.88% #20 11.12%  #5 10.88% #30 3.56%
\end{verbatim}
The first three columns are known from \texttt{convolute}.
The next columns (the switch \texttt{-{}-limit 5} limits the output to five columns) show first the channel index and then the relative size of the corresponding contribution.
Since the contribution of a partonic channel can be negative, the columns are sorted ignoring the sign of the contribution.
The first line shows that for bin 0, i.e.\ for the range $\SI{116}{\giga\electronvolt} < M_{\ell \bar{\ell}} < \SI{130}{\giga\electronvolt}$, the cross section is dominated by partonic channel \texttt{\#15} (\SI{27.42}{\percent}), following by partonic channel \texttt{\#0} with same size, then channel \texttt{\#5}, etc.
The meaning of the channel numbers is given by using the subcommand \texttt{luminosity} (only an excerpt is shown):
\begin{verbatim}
id    entry
--+------------+------------
0  1 × ( 2, -2) 1 × ( 4, -4)
5  1 × ( 1, -1) 1 × ( 3, -3)
15 1 × (-4,  4) 1 × (-2,  2)
20 1 × (-3,  3) 1 × (-1,  1)
30 1 × (22, 22)
\end{verbatim}
This shows that channel \texttt{\#0} represents the up-type quark--anti-quark contributions (shown with PDG id 2 and 4 for up and charm quarks, which have the same matrix elements), channel \texttt{\#15} is the same channel with its initial states transposed, channels \texttt{\#5} and \texttt{\#20} are the down-type quark--anti-quark channels, and channel \texttt{\#30} is the photon--photon channel.
The size of the remaining channels is smaller than the photon--photon channel.
The factors \texttt{1} are not important here, but in general they can contain CKM values and charge factors that, if kept in the squared matrix elements, would not allow for sharing a single matrix element for different quark flavours and therefore slow down the calculation.
A complete list of all channels and of their contribution to the cross section for all of the processes discussed in section~\ref{sec:results} is collected in appendix~\ref{app:lumis}.

@@

\subsection{Example Monte Carlo program in C++}
\label{app:example-program}

@@

The following listing shows how to setup \textsc{PineAPPL} using its C interface in a simple Monte Carlo integrator for calculating the double-photon contribution to Drell--Yan lepton-pair production at the LHC.
All \textsc{PineAPPL} functions have the prefix \texttt{pineappl\_}.
The full example together with a makefile can be found at \url{https://github.com/N3PDF/pineappl/tree/master/examples/capi-dy-aa}.
The documentation of the C API can be found at \url{https://docs.rs/pineappl_capi}.

\begin{lstlisting}[language=C++,mathescape=true]
#include <pineappl_capi.h>
#include <LHAPDF/LHAPDF.h>

#include <cmath>
#include <cstddef>
#include <cstdio>
#include <random>
#include <vector>

double int_photo(double s, double t, double u) {
    double alpha0 = 1.0 / 137.03599911;
    return alpha0 * alpha0 / 2.0 / s * (t / u + u / t);
}

struct Psp2to2 {
    double s;
    double t;
    double u;
    double x1;
    double x2;
    double jacobian;
};

Psp2to2 hadronic_pspgen(std::mt19937& rng, double mmin, double mmax) {
    using std::acos;
    using std::log;
    using std::pow;

    double smin = mmin * mmin;
    double smax = mmax * mmax;

    double r1 = std::generate_canonical<double, 53>(rng);
    double r2 = std::generate_canonical<double, 53>(rng);
    double r3 = std::generate_canonical<double, 53>(rng);

    double tau0 = smin / smax;
    double tau = pow(tau0, r1);
    double y = pow(tau, 1.0 - r2);
    double x1 = y;
    double x2 = tau / y;
    double s = tau * smax;

    double jacobian = tau * log(tau0) * log(tau0) * r1;

    // theta integration (in the CMS)
    double  cos_theta = 2.0 * r3 - 1.0;
    jacobian *= 2.0;

    double  t = -0.5 * s * (1.0 - cos_theta);
    double  u = -0.5 * s * (1.0 + cos_theta);

    // phi integration
    jacobian *= 2.0 * acos(-1.0);

    return { s, t, u, x1, x2, jacobian };
}

void fill_grid(pineappl_grid* grid, std::size_t calls) {
    using std::acosh;
    using std::fabs;
    using std::log;
    using std::sqrt;

    auto rng = std::mt19937();

    // in GeV^2 pbarn
    double  hbarc2 = 389379372.1;

    for (std::size_t i = 0; i != calls; ++i) {
        // generate a phase-space point
        auto tmp = hadronic_pspgen(rng, 10.0, 7000.0);
        auto s = tmp.s;
        auto t = tmp.t;
        auto u = tmp.u;
        auto x1 = tmp.x1;
        auto x2 = tmp.x2;
        auto jacobian = tmp.jacobian;

        double ptl = sqrt((t * u / s));
        double mll = sqrt(s);
        double yll = 0.5 * log(x1 / x2);
        double ylp = fabs(yll + acosh(0.5 * mll / ptl));
        double ylm = fabs(yll - acosh(0.5 * mll / ptl));

        jacobian *= hbarc2 / calls;

        // cuts for LO for the invariant-mass slice containing the
        // Z-peak from CMSDY2D11
        if ((ptl < 14.0) || (fabs(yll) > 2.4) || (ylp > 2.4)
            || (ylm > 2.4) || (mll < 60.0) || (mll > 120.0))
        {
            continue;
        }

        auto weight = jacobian * int_photo(s, u, t);
        double  q2 = 90.0 * 90.0;

        pineappl_grid_fill(grid, x1, x2, q2, 0, fabs(yll), 0, weight);
    }
}

int main() {
    // create a new luminosity function for the $\gamma\gamma$ initial state
    auto* lumi = pineappl_lumi_new();
    int32_t pdg_ids[] = { 22, 22 };
    double ckm_factors[] = { 1.0 };
    pineappl_lumi_add(lumi, 1, pdg_ids, ckm_factors);

@@

    // destroy the object
    pineappl_grid_delete(grid);
}
\end{lstlisting}

Updated installation instructions are kept in the file \texttt{README.md} in \textsc{PineAPPL}'s repository at \url{https://github.com/N3PDF/pineappl/} and on its homepage \url{https://n3pdf.github.io/pineappl/}.

After being installed, one can compile and link against the library.
See appendix~\ref{app:example-program} for an example.

### 2020-08-21 paper/sec-pineappl.tex
adding sentence about the python module
Improve PineAPPL appendix
- move installation appendix towards the end
- include complete code example

\item \textsc{PineAPPL} offers an easy-to-use interface written in the C programming language, that allows MC integrators to read and write \textsc{PineAPPL} grids.
C was chosen because it can be easily interfaced with both Fortran and C++, the two main programming languages in which most MC integrators are written.
The interface consists of roughly thirty functions, among which only a handful are needed in practice.
See appendix~\ref{app:example-program} for an example.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we have chosen a static scale, so that we do not need to interpolate in the $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them\footnote{The technical term for this interpolation problem is sometimes called \emph{subtabulation}.}.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u_1(y_1(x_1)),u_2(y_2(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula:
\begin{equation}
a_{i,j} = a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2)
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting from the smallest index in the square, $k_i$ and $k_j$.
Finally, we remapped
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part gives the location between the nearest grid points.

>>>>

\item \textsc{PineAPPL} offers an easy-to-use interface written in the C programming language, that allows MC integrators to read and write \textsc{PineAPPL} grids.
C was chosen because it can be easily interfaced with both Fortran and C++, the two main programming languages in which most MC integrators are written.
The interface consists of roughly thirty functions, among which only a handful are needed in practice.
We also provide and support a Python package based on this C interface.
See appendix~\ref{app:example-program} for an example.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we have chosen a static scale, so that we do not need to interpolate in the $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them\footnote{The technical term for this interpolation problem is sometimes called \emph{subtabulation}.}.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u(y(x_1)),u(y(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula:
\begin{equation}
a_{i,j} = a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2)
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting from the smallest index in the square, $k_i$ and $k_j$.
Finally, we remapped
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part gives the location between the nearest grid points.
